Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-10-29 ms/ms.tex
revised ms sent to coauthors

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi

@@

\(^2\) School of Life Sciences, University of California, Davis, CA, USA
95616

@@

\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and that leaves with more
similar conductances at each surface benefit more than leaves with few
stomata on the upper surface.

@@

\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, leaf, \emph{Sida
fallax}, stomata

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts, yet paradoxically
this seemingly simple adaptation is uncommon in nature and we don't know
why. Understanding variation in stomatal traits like amphistomy is
imperative because these tiny pores play an outsized ecological role in
the global carbon and water cycles (Hetherington and Woodward, 2003;
Berry et al., 2010). A widely applicable, accurate representation of how
stomata mediate the relationship between CO\(_2\) gained through
photosynthesis and water lost through evapotranspiration is essential to
predict future climate using Earth Systems Models (Jarvis, 1976; Ball et
al., 1987; Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997).
Optimality models accurately predict the major cause of water loss,
stomatal conductance (\(g_\mathrm{sw}\)), by assuming plants maximize
carbon gain minus a cost of water (Cowan and Farquhar, 1977; Givnish,
1986; Medlyn et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks
et al., 2018; Deans et al., 2020; Franklin et al., 2020; Wang et al.,
2020; Harrison et al., 2021). Despite the success of optimality modeling
in predicting \(g_\mathrm{sw}\), the same modeling approach has so far
failed to explain the rarity of amphistomatous leaves (Muir, 2019).
\textbf{This gap between theory and observations strongly implies that
we remain ignorant about some key benefits and costs associated with
stomata.}

@@

Amphistomatous leaves should be common because, all else being equal, a
leaf with a given number of stomata per area could increase its
photosynthetic rate simply by apportioning approximately half its
stomata to each surface (Parkhurst, 1978; Gutschick, 1984a, b). The key
difference between a hypo- and amphistomatous leaf, holding all other
factors constant, is that an amphistomatous leaf has two parallel
diffusion paths through the internal airspace to any given chloroplast.
Those airspaces pose a resistance for CO\(_2\) diffusion, so CO\(_2\)
concentration drops as it approaches chloroplasts. Shorter pathways mean
a smaller drop in CO\(_2\) concentration. Thus, chloroplasts in
amphistomatous leaves experience higher CO\(_2\) concentrations than in
hypostomatous leaves, thereby increasing photosynthesis. The airspace
resistance (or its inverse, the airspace conductance,
\(g_\mathrm{ias}\)) is rarely measured directly and there is
disagreement between empirical (Parkhurst and Mott, 1990; Morison et
al., 2005; Evans et al., 2009; Tomás et al., 2013; Earles et al., 2018;
Šantrůček et al., 2019; Nobel, 2020; Harwood et al., 2021; Márquez et
al., 2023) and theoretical models (Tholen and Zhu, 2011; Ho et al.,
2016; Théroux-Rancourt et al., 2021). The \(g_\mathrm{ias}\) in thin,
porous leaves may be so large as to be inconsequential given much lower
conductances for other components of the diffusion pathway, whereas the
\(g_\mathrm{ias}\) of thick leaves with little airspace may greatly
hinder CO\(_2\) diffusion to chloroplasts. Amphistomy should confer the
largest photosynthetic benefit in leaves with intrinsically low
\(g_\mathrm{ias}\).

Amphistomy should also enhance photosynthesis when leaf boundary layer
resistance is high, because apportioning total flux between two boundary
layers rather than one results in a smaller CO\(_2\) concentration drop
between the atmosphere and stomata. A similar effect has been validated
with a computer model and measurements for transpiration: amphistomatous
leaves lose somewhat more water for the same vapor pressure deficit and
total \(g_\mathrm{sw}\) (Foster and Smith, 1986), but the additional
carbon gain should be enough to offset this cost under most realistic
conditions (Muir, 2019).

@@

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), resulting in lower \(g_\mathrm{ias}\)
(Parkhurst, 1978). A nonmutually exclusive hypothesis is that if sun
leaves have a stomatal density ratio closer to 0.5 (same density on each
leaf surface), this will confer a greater advantage than an
amphistomatous leaf with most stomata on one surface. In other words,
amphistomy doesn't make much difference if one leaf surface has few open
stomata on it. We therefore predict that sun leaves will have greater
\(\mathrm{AA}\) possibly because they have thicker leaves and/or
\(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that it accounts for differences in both stomatal density and size
between surfaces.

@@

\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}

@@

\hypertarget{plant-sampling}{%
\subsection{Plant sampling ---}\label{plant-sampling}}

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting Yorkston and Daehler (2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
during one three-day trip. We haphazardly sampled eight plants
distributed evenly between the highest and lowest elevation plants along
a transect at each site. For safety and conservation reasons, transects
were along a trail or road. We did not sample small individuals if there
was risk removing leaves would cause mortality. From each plant, we
collected two fully expanded leaves for trait measurements. We sampled
stomatal traits on all leaves; leaf thickness on one leaf from three
randomly selected plants per site; and, due to limited time, a single
leaf from a single plant at the middle of each transect for gas exchange
measurements.

\hypertarget{tbl-locations}{}
\begin{table}
\caption{\label{tbl-locations}ʻIlima study sites for this study. }\tabularnewline

@@

We estimated the stomatal density and size on ab- and adaxial leaf
surfaces from all leaves. For pubescent leaves (usually coastal), we
dried and pressed leaves for \(\approx 1\) week (Hill et al., 2014),
carefully scraped trichomes off with a razor blade, and rehydrated the
leaf. For glabrous leaves, we used fresh leaves. We applied clear nail
polish to both leaf surfaces of fresh or rehydrated leaves in the middle
of the lamina away from major veins. After nail polish dried, we mounted
impressions on a microscope slide using transparent tape (Mott and
Michaelson, 1991). We digitized a portion of each leaf surface
impression using a brightfield microscope (Leica DM2000, Wetzlar,
Germany). We counted all stomata and divided by the visible leaf area
(0.890 mm\(^2\)) to estimate density and measured guard cell length from
five randomly chosen stomata per field using ImageJ (Schneider et al.,
2012).

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). We also estimated
`amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu} Parkhurst (1978),
but with modification. For each leaf, we measured the photosynthetic
rate of an untreated amphistomatous leaf (\(A_{\mathrm{amphi}}\)) over a
range of \(g_\text{sw}\) values. We refer to this as an
\(A \textendash g_\text{sw}\) curve. We compared the
\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
which is the same leaf but with gas exchange through the upper surface
blocked by a neutral density plastic (propafilm). Hypostomy refers to
leaves with stomata only present on the lower, typically abaxial,
surface. We refer to the untreated and partially blocked leaves as
``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
calculated as the log-response ratio of \(A\) compared at the same total
\(g_\text{sw}\):

@@

The log-response ratio is easy to interpret because values above 0
indicate a photosynthetic advantage of amphistomy, whereas values less
than 0 indicate a disadvantage. The response ratio is also better than
the absolute difference because it indicates a proportional change in
\(A\), which facilitates comparisons across leaves and environments with
different baseline photosynthetic rates. The irradiance of the light
source in the pseudohypo leaf is higher because the propafilm reduces
transmission. Because the propafilm reduces transmission, we used a
higher incident \(\mathrm{PPFD}\) for pseudohypo leaves (higher by a
factor 1/0.91, the inverse of the measured transmissivity of the
propafilm). We also set the stomatal conductance ratio, for purposes of
calculating boundary layer conductance, to 0 for pseudohypo leaves
following manufacturer directions.

To control for order effects, we alternated between starting with amphi
(untreated) or pseudohypo (treated) leaf measurements. In the field, we
acclimated the focal leaf to high light
(\(\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
high relative humidity (\(\mathrm{RH} = 70\%\)),
\(T_\text{leaf} = 25-30~° \text{C}\), depending on ambient temperature,
until \(A\) and \(g_\text{sw}\) reach their maximum. We used these data
as our estimates of maximum \(A\) and \(g_\text{sw}\). After that, we
decreased \(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal
closure without photosynthetic downregulation. All other environmental
conditions in the leaf chamber remained the same. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process with the
other treatment (amphi or pseudohypo).

@@

It is not feasible to record \(A_\mathrm{amphi}\) and
\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
we fit \(A \textendash g_\text{sw}\) curves using a linear regression of
\(\text{log}(g_\mathrm{sw})\) on \(A\) to interpolate modeled \(A\) for
amphi and pseudohypo leaves at the same \(g_\text{sw}\). In 10 of 12
leaves, the minimum \(g_\text{sw}\) of the amphi curve was smaller than
the maximum \(g_\text{sw}\) of the pseudohypo curve (i.e.~the curves
overlapped for a range of \(g_\text{sw}\) values). In those cases, we
estimated \(\mathrm{AA}\) at the \(g_\mathrm{sw}\) value in the middle
of the range of overlap. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. We fit Bayesian regressions using the \emph{R} package
\textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling in
\emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations. We estimated parameters and confidence intervals
as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\).

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations. We estimated the relationship between
population average leaf thickness and \(\mathrm{AA}\) measured from a
single individual per population. We used this approach most of the
variation in leaf thickness was among sites and the plant selected for
gas exchange measurements was not always among the plants randomly
selected for leaf thickness, precluding individual level correlation. We
propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness by
integrating over the entire posterior distribution sample for each
variable. The key prediction is that the effect of leaf thickness on
\(\mathrm{AA}\) is positive.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations. From each
posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\) as:

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\(^2\) Department of Plant Sciences, University of California, Davis,
CA, USA 95616

@@

\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because 1) they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and 2) that they benefit more
because they have similar conductance on each surface, as opposed to
most of the conductance being on the lower (abaxial) surface.

@@

\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, \emph{Sida fallax},
stomata

@@

Amphistomy, the presence of stomata on both lower and upper surfaces of
broad leaves, should increase carbon gain by reducing the average
diffusion pathlength between stomata and chloroplasts, yet paradoxically
this seemingly simple adaptation is uncommon in nature and we don't know
why. Understanding variation in stomatal traits like amphistomy is
imperative because these tiny pores play an outsized ecological role in
the global carbon and water cycles (Hetherington and Woodward, 2003;
Berry et al., 2010). A widely applicable, accurate representation of how
stomata mediate the relationship between CO\(_2\) gained through
photosynthesis and water lost through transpiration is essential to
predict future climate using Earth Systems Models (Jarvis, 1976; Ball et
al., 1987; Collatz et al., 1991; Leuning, 1995; Sellers et al., 1997).
Optimality models accurately predict the major cause of water loss,
stomatal conductance (\(g_\mathrm{sw}\)), by assuming plants maximize
carbon gain minus a cost of water (Cowan and Farquhar, 1977; Givnish,
1986; Medlyn et al., 2011; Lin et al., 2015; Wang et al., 2017; Franks
et al., 2018; Deans et al., 2020; Franklin et al., 2020; Wang et al.,
2020; Harrison et al., 2021). Despite the success of optimality modeling
in predicting \(g_\mathrm{sw}\), the same modeling approach has so far
failed to explain the rarity of amphistomatous leaves (Muir, 2019).
\textbf{This gap between theory and observations strongly implies that
we remain ignorant about some key benefits and costs associated with
stomata.}

@@

Amphistomatous leaves should be common because, all else being equal, a
leaf with a given number of stomata per area could increase its
photosynthetic rate simply by apportioning approximately half its
stomata to each surface (Parkhurst, 1978; Gutschick, 1984a, b). The key
difference between a hypo- and amphistomatous leaf, holding all other
factors constant, is that an amphistomatous leaf has two parallel
diffusion paths through the internal airspace to any given chloroplast.
Those airspaces pose a resistance for CO\(_2\) diffusion, so CO\(_2\)
concentration drops as it approaches chloroplasts. Shorter pathways mean
a smaller drop in CO\(_2\) concentration. Thus, chloroplasts in
amphistomatous leaves experience higher CO\(_2\) concentrations than in
hypostomatous leaves, thereby increasing photosynthesis. The airspace
resistance (or its inverse, the airspace conductance,
\(g_\mathrm{ias}\)) is rarely measured directly and there is
disagreement between empirical (Parkhurst and Mott, 1990; Morison et
al., 2005; Evans et al., 2009; Tomás et al., 2013; Earles et al., 2018;
Šantrůček et al., 2019; Nobel, 2020; Harwood et al., 2021; Márquez et
al., 2023) and theoretical models (Tholen and Zhu, 2011; Ho et al.,
2016; Théroux-Rancourt et al., 2021). The \(g_\mathrm{ias}\) in thin,
porous leaves may be so large as to be inconsequential given much lower
conductances for other components of the diffusion pathway, whereas the
\(g_\mathrm{ias}\) of thick leaves with little airspace may greatly
hinder CO\(_2\) diffusion to chloroplasts. Amphistomy should confer the
largest photosynthetic benefit in leaves with intrinsically low
\(g_\mathrm{ias}\). The airspace conductance is one component of the
overall mesophyll conductance, \(g_\mathrm{m}\), which is often strongly
influenced by the chloroplast surface area exposed to airspace and
mesophyll cell wall thickness (Evans et al., 2009; Gago et al., 2020;
Flexas et al., 2021). Hence, thicker leaves may compensate for lower
\(g_\mathrm{ias}\) through increased chloroplast surface area exposed to
airspace (Terashima et al., 2006), but will still benefit from
amphistomy as long as \(g_\mathrm{ias}\) is finite.

Amphistomy should also enhance photosynthesis when leaf boundary layer
resistance is high, because apportioning total flux between two boundary
layers rather than one results in a smaller CO\(_2\) concentration drop
between the atmosphere and stomata. A similar effect has been validated
with a computer model and measurements for transpiration: amphistomatous
leaves lose somewhat more water for the same vapor pressure deficit and
total \(g_\mathrm{sw}\) (Foster and Smith, 1986), but the additional
carbon gain should be enough to offset this cost under most realistic
conditions (Muir, 2019). However, if minimal stomatal conductance is
related to stomatal density (Drake et al., 2013; Márquez et al., 2022)
and the upper boundary layer conductance is higher, then amphistomy
could cause additional, unavoidable water loss.

@@

The association between high light and amphistomy suggests that `sun'
leaves have the most to gain in terms of increased photosynthesis from
having stomata on both surfaces, as Mott et al. (1982) hypothesized.
Parkhurst (1978) proposed quantifying this benefit as `amphistomy
advantage' (\(\mathrm{AA}\)), which we adopt here with some modification
(see Materials and Methods). This hypothesis has never been tested
directly by comparing the photosynthetic rate of an amphistomatous leaf
to that of an otherwise identical hypostomatous leaf with the same total
stomatal conductance under the same conditions. We propose a
straightforward method to do this by experimentally creating a
pseudohypostomatous leaf with gas exchange blocked through the upper
surface (see Materials and Methods). One reason that sun leaves might
have greater \(\mathrm{AA}\) is that they are usually thicker or denser
(Poorter et al., 2019), which will often result in lower
\(g_\mathrm{ias}\) either by increasing the diffusion path length
(Parkhurst, 1978) or making the airspace less porous. A nonmutually
exclusive hypothesis is that if sun leaves have a stomatal density ratio
closer to 0.5 (same density on each leaf surface), this will confer a
greater advantage than an amphistomatous leaf with most stomata on one
surface. In other words, amphistomy doesn't make much difference if one
leaf surface has few open stomata on it. We therefore predict that sun
leaves will have greater \(\mathrm{AA}\) possibly because they have
thicker leaves and/or \(\mathrm{SR}\) closer to 0.5. We actually report
\(g_{\mathrm{smax,ratio}}\), which is similar to \(\mathrm{SR}\) except
that it accounts for differences in both stomatal density and size
between surfaces.

@@

\begin{figure}[H]
  \includegraphics[height=3.25in]{../figures/study-system.pdf}
  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

@@

\hypertarget{plant-sampling}{%
\subsection{Plant sampling and climate ---}\label{plant-sampling}}

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on
Hawaiʻi Island by consulting Yorkston and Daehler (2006) and citizen
scientist records on iNaturalist (Anon, 2022)
(\autoref{fig:study-system}c; \autoref{tbl-locations}). We avoided sites
that appeared to be cultivated. We visited sites between August and
November 2022. For logistical reasons, the sites on Hawaiʻi were sampled
during one three-day trip. We haphazardly sampled eight plants
distributed evenly between the highest and lowest elevation plants along
a transect at each site. For safety and conservation reasons, transects
were along a trail or road. We did not sample small individuals if there
was risk removing leaves would cause mortality. From each plant, we
collected two fully expanded leaves for trait measurements. We sampled
stomatal traits on all leaves; leaf thickness on one leaf from three
randomly selected plants per site; and, due to limited time, a single
leaf from a single plant at the middle of each transect for gas exchange
measurements. We downloaded climatic data on mean annual temperature,
solar radiation, and vegetation height from the Climate and Solar
Radiation of Hawaiʻi databases (Giambelluca et al., 2014) using the
latitude and longitude at the middle of each transect. We also
downloaded mean annual precipitation from 1978-2007 from the Rainfall
Atlas of Hawaiʻi (Giambelluca et al., 2013). The spatial resolution is
approximately \(234 \times 250\)m. The temperature data are calibrated
from networks of meteorological stations operating in the late 20th
century and 21st century; the solar radiation data are calibrated from
satellite measurements collected between 2002 and 2009 (Giambelluca et
al., 2014). We tested whether climatic variables differed among our
coastal and montane populations using Welch's two-sample \(t\)-test.

\hypertarget{tbl-locations}{}
\begin{table}
\caption{\label{tbl-locations}ʻIlima study site location information. }\tabularnewline

@@

We estimated the stomatal density and size on ab- and adaxial leaf
surfaces from all leaves. For pubescent leaves (usually coastal), we
dried and pressed leaves for \(\approx 1\) week (Hill et al., 2014),
carefully scraped trichomes off with a razor blade, and rehydrated the
leaf. Rehydration restores leaf area to its fresh value (Blonder et al.,
2012). For glabrous leaves, we used fresh leaves. We applied clear nail
polish to both leaf surfaces of fresh or rehydrated leaves in the middle
of the lamina away from major veins. After nail polish dried, we mounted
impressions on a microscope slide using transparent tape (Mott and
Michaelson, 1991). We digitized a portion of each leaf surface
impression using a brightfield microscope (Leica DM2000, Wetzlar,
Germany). We counted all stomata and divided by the visible leaf area
(0.890 mm\(^2\)) to estimate density and measured guard cell length from
five randomly chosen stomata per field using ImageJ (Schneider et al.,
2012).

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). The midday
irradiance in coastal ʻilima typically meets or even exceeds a PPFD of
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) and previous
experiments with sun leaves revealed that
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) is always at or near
saturating irradiance. Even though lower irradiance may be saturating
for montane leaves, we used this higher value for all leaves to
standardize conditions.

We also estimated `amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu}
Parkhurst (1978), but with modification. For each leaf, we measured the
photosynthetic rate of an untreated amphistomatous leaf
(\(A_{\mathrm{amphi}}\)) over a range of \(g_\text{sw}\) values. We
refer to this as an \(A \textendash g_\text{sw}\) curve, which is
described in more detail below. We compared the
\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
which is the same leaf but with gas exchange through the upper surface
blocked by a neutral density plastic (propafilm). Hypostomy refers to
leaves with stomata only present on the lower, typically abaxial,
surface. We refer to the untreated and partially blocked leaves as
``amphi'' and ``pseudohypo'', respectively. \(\mathrm{AA}\) is
calculated as the log-response ratio of \(A\) compared at the same total
\(g_\text{sw}\):

@@

The log-response ratio is commonly used social and biological sciences
(e.g. Hedges et al. (1999)). It is straightforward to interpret because
values above 0 indicate a photosynthetic advantage of amphistomy,
whereas values less than 0 indicate a disadvantage. The log-response
ratio is preferable to the absolute difference because it indicates a
proportional change in \(A\), which facilitates comparisons across
leaves and environments with different baseline photosynthetic rates.
The irradiance of the light source in the pseudohypo leaf was higher
because the propafilm reduces transmission. To compensate for reduced
transmission, we increased incident \(\mathrm{PPFD}\) for pseudohypo
leaves by a factor 1/0.91, the inverse of the measured transmissivity of
the propafilm. We also set the stomatal conductance ratio, for purposes
of calculating boundary layer conductance, to 0 for pseudohypo leaves
following manufacturer directions.

\autoref{fig:ags-curve} illustrates our method for collecting
\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
an amphi (untreated) curve and a pseudohypo (treated) curve. To control
for order effects, we alternated between starting with amphi or
pseudohypo leaf measurements. In the field, we acclimated the focal leaf
to saturating light and high relative humidity (\(\mathrm{RH} = 70\%\)),
as described above, until \(A\) and \(g_\text{sw}\) reach their maximum.
We used these data as our estimates of maximum \(A\) and
\(g_\text{sw}\). After that, we decreased \(\mathrm{RH}\) to
\(\approx 10\%\) to induce rapid stomatal closure without biochemical
downregulation. Hence, \(A_\text{amphi}\) and \(A_\text{hypo}\) were
both measured at low chamber humidity after the leaf had acclimated to
high humidity. All other environmental conditions in the leaf chamber
remained the same. We logged data until \(g_\text{sw}\) reached its
nadir. We then repeated the process of acclimating the leaf to 70\%
\(\mathrm{RH}\) and inducing stomatal closure with low \(\mathrm{RH}\)
with the other treatment (amphi or pseudohypo). We did not detect an
effect of treatment order on \(\mathrm{AA}\) (results not shown).

@@

It is not feasible to record \(A_\mathrm{amphi}\) and
\(A_\mathrm{hypo}\) at the exact same \(g_\text{sw}\). To overcome this,
we fit \(A \textendash g_\text{sw}\) curves using a linear regression of
\(\text{log}(g_\mathrm{sw})\) on \(A\) to interpolate modeled \(A\) for
amphi and pseudohypo leaves at the same \(g_\text{sw}\). Let
\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) be the estimated
\(A\) of the amphi and pseudohypo leaves, respectively. We estimated
these quantities at the same \(g_\mathrm{sw}\) using fitted parameters
(\(\hat{\beta}\)'s):

\[\hat{A}_\text{amphi} = \hat{\beta}_{0,\text{amphi}} + \hat{\beta}_{1,\text{amphi}} \times \text{log}(g_\mathrm{sw})\]
\[\hat{A}_\text{hypo} = \hat{\beta}_{0,\text{hypo}} + \hat{\beta}_{1,\text{hypo}} \times \text{log}(g_\mathrm{sw})\]

In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
those cases, we estimated \(\hat{A}_\text{amphi}\) and
\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
the range of overlap between the curves. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. The vertical lines in \autoref{fig:licor} show the
\(g_\text{sw}\) for each leaf. \(\mathrm{AA}\) for each leaf was
calculated \(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) using
the log-response ratio shown above.

To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
curve for each leaf, we fit Bayesian regressions using the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling
in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated parameters and confidence
intervals as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
95\% confidence intervals of
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations per chain. We estimated the
relationship between population average leaf thickness and
\(\mathrm{AA}\) measured from a single individual per population. We
used this approach because most of the variation in leaf thickness was
among sites and the plant selected for gas exchange measurements was not
always among the plants randomly selected for leaf thickness, precluding
individual level correlation. We propagated uncertainty about in
\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
entire posterior distribution sample for each variable. The key
prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
positive, meaning the 95\% confidence interval of the slope should be
positive and not encompass 0.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations per chain.
From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
as:

\[g_\mathrm{smax,ratio} = \frac{g_\mathrm{smax,upper}}{g_\mathrm{smax,lower} + g_\mathrm{smax,upper}},\]
